---
title: "Using the `here` package for reproducible workflows"
date: 2024-03-17
description: "Gentle reminders when using the `here` package for your `RStudio` projects"
categories:
 - r
 - packages
 - here
 - working dir
 - setwd
 - programming
 - code errors
 - workflow
 - reproducibility
format:
  html:
    code-fold: true
    code-tools: true
#doi: 10.59350/bje88-8r592
citation: true
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.retina = 3,
                      fig.width = 5.5, fig.height = (5.5 * 0.618),
                      out.width = "90%", collapse = TRUE)
options(digits = 3, width = 300)
```

## TL; DR: 

Use the `here` package instead of `setwd()` or `getwd()` to increase reproducibility and avoid wasting your and other people's time.

Since I created this website, I've been coding, writing, and reading a lot more which has unequivocally led to a mountain of new files and  the forging of new paths– quite literally:smile:. At first, I kept things pretty organized, but now it's nearly impossible to know where I saved `such_and_such.txt` file without wasting at least 5 minutes of my day. 

This is what I used to do:

```{r, eval = FALSE echo = TRUE}
url1 <- "https://somefile_online_data_source_here.com"
download.file(url1, destfile = "./data_file_here.zip")
unzip("data_file_here.zip", exdir = getwd())
Dat <- readRDS("summaryDat.rds")
Dat2 <- readRDS("SummaryDat2.rds")
```

This is fine until it's time to either save or get new files. My directory will be anywhere on my device unless I have previously specified it using `setwd()`, but this strategy will soon be an obstacle to saving new information in an organized and reproducible way. If later on I change my `R` scripts to a different folder the original file path won't work anymore. 

## `here` comes the package :file_folder:

The `here` package allows you to set up a relative path mapped onto your R project directory on every device regardless of your absolute path directory.

### the `here` function 

Here you can see what my root directory is and how I'm going up one folder with each iteration of the `here` command.

```{r, echo = TRUE}
here::here()
# [1] "C:/Users/jpmonteagudo/Desktop/R/Project"
  here::here("blog")
# [1] "C:/Users/jpmonteagudo/Desktop/R/Project/blog"
  here::here("2024")
# [1] "C:/Users/jpmonteagudo/Desktop/R/Project/2024"
  here::here("02")
# [1] "C:/Users/jpmonteagudo/Desktop/R/Project/02"
  here::here("post")
# [1] "C:/Users/jpmonteagudo/Desktop/R/Project/post"
  here::here("scripts")
# [1] "C:/Users/jpmonteagudo/Desktop/R/Project/scripts"
```

```{r echo = TRUE}
here::here("blog","2024","02","post","scripts","Dat.R")
# [1] "C:/Users/jpmonteagudo/Desktop/R/Poroject/blog/2024/02/post/scripts/Dat.R"
```

I can also go up several folders by calling the full relative path in one single command. However, when I call the `here` function again, it sends me back to my root directory.

```{r echo = TRUE}
here::here()
# [1] "C:/Users/jpmonteagudo/Desktop/R/Project"
```

It looks like I would use the `here` function to get or write files and not just be there. If I don't need anything from my subdirectory, then `R` will go back to where the `.Rproj` file is located. For example, if I want to save a `.png` file with a series of plots to one of my folders, I can use the `here::here` function to assign the relative path within my save command.

```{r, eval = FALSE echo = TRUE}
ggsave("plots.png",arranged_plots, 
       path = here::here("blog","2024","02","post","img"),
                width = 800,
                height = 600,
                units = "px",
                dpi = 72)
```

### the `set_here` function 

Another function available in this package is the `set_here` function which creates a `.here` file anywhere in your project so you can use the directory that contains this file as `root`. This is the description in the function's syntax

>When `here` encounters such a file, it uses the directory that contains this file as root. This is useful if none of the default criteria apply. You need to restart the R session so that `here()` picks up the newly created file.

```{r, eval = FALSE echo = TRUE}
here::set_here("blog/2024")
# Created file .here in C:\Users\jpmonteagudo\Desktop\R\Project\blog\2024. Please start a new R session in the new project directory.
```

Next, I start a new `R` session from this new directory, and `RStudio` will automatically set my directory to this folder. I don't need to open the `R` project to reach this new directory. It will give me access to the folder's files and I can then set a relative path to other folders and files.

```{r, eval = FALSE echo = TRUE}
here::here()
# [1] "C:/Users/jpmonteagudo/Desktop/R/Project/blog/2024"
```

From this new directory, I can reach files anywhere by just calling `here::here()` function to specify the location of the file I'd like to get. 

### The confusing `i_am` function

