---
title: "A parametric quantile beta regression for modeling case fatality rates of COVID-19. Building models using the betareg and RBE3 packages"
date: 2024-02-26
description: "Using the betareg and RBE3 packages to model ratios and proportions using COVID-19 simulated data"
categories:
  - r
  - betareg
  - percentile
  - regression
  - beta distribution
  - COVID-19
  - GLMs
  - modeling
resources:
  - "_book/**"
  - "img/*"
doi: 10.59350/qgdjk-8ph86
citation: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.retina = 3,
                      fig.width = 5.5, fig.height = (5.5 * 0.618),
                      out.width = "90%", collapse = TRUE)
options(digits = 3, width = 300)
```

## Quantile Beta Regression

The quantile beta regression model discussed in the 2021 paper by

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
